---
title: 'Exploring the ToothGrowth Dataset'
author: "Robert Sizemore"
date: "9/4/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
```


In this document, we want to perform a basic exploratory data analysis on the ToothGrowth dataset. According to the documentation, this dataset looks at the length of the odontoblasts (cells responsible for tooth growth) in 60 guinea pigs which were given varying doses (0.5, 1, and 2 mg/day) of Vitamin C via two different delivery methods (OJ or VC). Before we try any analyses on the data, we take a quick look at the summary of the dataset. For convenience, we convert the dosage into a factor variable with 3 levels:

```{r}
library(ggplot2)
data("ToothGrowth")
data <- data.frame(len = ToothGrowth$len, 
                   supp = ToothGrowth$supp, dose = as.factor(ToothGrowth$dose))
summary(data)
```

We see that number of observations is divided evenly among each dose and delivery method and that the observed tooth lengths range from 4.2 to 33.9. Furthermore, the observations are cleanly divided among the administration methods and dosage amounts (30/20 each respectively).

## A Quick Look at the Data

The purpose of this data is to look at the effects of dosage and delivery methods on observed tooth growth, so before performing any tests, lets take a look at the data:

```{r, out.width = '50%'}
qplot(len, data = data, fill = supp, binwidth = 2.5, xlab = "length", ylab = "count")
```

It appears that the way that the dose is administered seems to have an effect. The juice doses tends to have larger values, while the vitamin doses have a large spread but cluster more at the lower values. We can do a similar thing with the dosages:

```{r, out.width = '50%'}
qplot(len, data = data, fill = dose, binwidth = 2.5, xlab = "length", ylab = "count")
```

This has a more pronounced visible separation with the higher doses corresponding to higher recorded lengths. We want to confirm this by running some statistical tests on our data.

## Comparing Method of Delivery 

We want to explore the effect of the means of delivery on the growth of odontoblast cells. Hence, the natural null hypotheses to test against is that changing the delivery method has no effect on tooth growth. Since we have no prior reason to suspect that one method is better than the other, thus we use the default two-sided t-test. Also, our data is not paired, so we explicit select paired = FALSE.


```{r}
suppt <- t.test(len ~ supp, data = data, paired = FALSE)
suppt
```

Recall that we are testing the null hypothesis that there is no difference in the means against the alternative hypothesis that the difference in the means is not zero. The calculated 95% confidence interval for the difference of the means is $(-0.17,7.57)$, which does contain $0$, so we cannot reject the null hypothesis. However, it is worth noting that our confidence interval barely contains $0$. In fact, the p-value for observing the difference in means for each group is $p = .06$, meaning there is only a 6% chance to see such a separation under the null hypothesis. It would be worthwhile to explore this further, as we only very narrowly fail to reject the null hypothesis, so the likelihood of a Type 2 error is fairly high.

## Comparing Dosage

Now we want to explore the effect of the dosage on the growth of odontoblast cells. The null hypothesis to test against is that increasing the dosage has no effect on tooth growth. First, we look at the growths for the low and mid-sized dosages to see if their means are likely to be different:

```{r}
lowmid <- subset(data, dose %in% as.factor(c(0.5,1)))
lowt <- t.test(len ~ dose, data = lowmid, paired = FALSE, var.equal = FALSE)
lowt
```

We see that the 95% confidence interval for the difference in the observed means for the low (0.5 mg/day) and middle (1 mg/day) dosages is (-11,-6) which does not contain 0. Furthermore, the p-value for observed the calculated difference of the mean under the null hypothesis is very small (~ $10^{-7}$), so we are pretty confident in our rejection of the null hypothesis. We do the same for the middle and high dosages:

```{r}
midhigh <- subset(data, dose %in% as.factor(c(1,2)))
midt <- t.test(len ~ dose, data = midhigh, paired = FALSE, var.equal = FALSE)
midt
```

We see that the 95% confidence interval for the difference in the observed means for the midsized (1 mg/day) and high (2 mg/day) dosages is (-9.0,-3.7) which does not contain 0. Furthermore, the p-value for observed the calculated difference of the mean under the null hypothesis is very small (~ $10^{-5}$), so again we are confident in our rejection of the null hypothesis. 

## Conclusions

We also looked at the ToothGrowth dataset which explores the effects of dosing Vitamin C on tooth growth in Guinea Pigs. This data looks at the size of the doses (0.5,1,2 mg/day) as well as the method of delivery (juice, vitamin supplement) in a total of 60 observations, hence a relatively small sample size.

For method of delivery, we were not able to reject the null hypothesis that there is no difference in the means using a 95% confidence interval. However, the p-value of the observed difference was very small, indicating a relatively high probability of a Type II error. 

When we compared the dosages, we calculated a large difference in the means. We used two T-tests to compare the small and medium dosages as well as the medium and high doses. In both cases we were able to confidently reject the null hypothesis that the difference in the means was zero. Hence, we would conclude that the mean growth is larger for medium doses than small doses and the mean growth is larger for large doses than medium doses.